\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}

### 2021-06-01 Manuscripts/Manuscript by chapters -v2/Full_document.tex
Updated with Discussion and conclusions

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, a \emph{linear mixed model} (LMEM). Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear. In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data. This paper summarizes the limitations of LMEMs and rm-ANOVA, presents the basic theory of GAMs, and demonstrates their implementation in \(\textsf{R}\) via the package \emph{mgcv} using simulated data that follows longitudinal trends reported in biomedical literature. To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:\_\_\_\_\_\_\_\_.

@@

The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.

A \emph{post hoc} analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.

@@

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}. Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}. And even more importantly, LMEMs also expect a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.

@@

In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively. Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently normally distributed with mean zero and variance \(\sigma^2_\mu\)).
In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy''). Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B). The linear models then can be expressed as

@@

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}. Dropout from patients and attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}. In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}. If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value. An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}.

In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.

This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and missing data. In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study. However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data. To better convey the issues of linearity and correlation in linear models fitted to non-linear data, simulation is used in the next section.

@@

To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \eqref{eq:linear-model}) is fitted to each group, using \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme}{[}\protect\hyperlink{ref-nlme}{49}{]}.
Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}, C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity.

Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect. In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group. From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D). A more comprehensive exploration of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.

@@

\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.}\label{fig:l-q-response}
\end{figure}

The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response},B). When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}, C. Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \ref{fig:l-q-response} A and B) . The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group. Fitting an rm-ANOVA model \eqref{eq:linear-model} to this data produces the fit that appears in panel D in Figure \ref{fig:l-q-response}.

A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\ref{fig:l-q-response}, D) indicates that the model is not capturing the changes within each group in a good way. Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model. Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare panels C and D). If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response. Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

@@

\hypertarget{gams-as-a-special-case-of-generalized-linear-models}{%
\section{GAMs as a special case of Generalized Linear Models}\label{gams-as-a-special-case-of-generalized-linear-models}}

@@

Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the function \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.

In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}. This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response. One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with the fit {[}\protect\hyperlink{ref-beck1998}{36}{]}. In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.

The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis functions} expansions of the covariates and by estimating random coefficients for these basis functions. A \emph{basis} is a set of functions that spans the space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}. For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions. In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.

A commonly used \emph{basis function} is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}. Cubic splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}. Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\ref{fig:l-q-response}, C). The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different basis function will be used. Because there are six timepoints for this group, five knots can be used. The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept). The choice of basis functions is already optimized in the package \emph{mgcv} depending on the number of knots. In Panel A of Figure \ref{fig:basis-plot}, the four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the ``wigliness'' of the fit, each of the basis functions of Panel A is penalized by multiplying it by a coefficient according to the penalty matrix of Panel B. The penalty reduces the ``wigliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

@@

\caption{Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Penalty matrix for the basis functions. Each basis function is penalized by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Penalized basis functions. Each of the four basis functions of panel A has been penalized by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, simulated values for the group appear as points.}\label{fig:basis-plot}
\end{figure}

@@

The previous sections provided the basic theoretical framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to LMEMs or rm-ANOVA. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}. The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.

@@

The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3(C) in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}. In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}, A and the inlet, respectively.

@@

An interaction model is typically the main interest in longitudinal biomedical data, as it takes into account the effect of treatment, time, and their combination. In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time. Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.

>>>>

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear. In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs, presents the basic theory of GAMs, and uses simulated data that follows trends reported in the biomedical literature to demonstrate their implementation in \(\textsf{R}\) via the package \emph{mgcv}. To make this work reproducible, the code and data used in this paper are available at:\_\_\_\_\_\_\_\_.

@@

The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.

@@

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}. Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}. And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.

@@

In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively. Independent errors \(\varepsilon_{tij}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy''). Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B). The linear models then can be expressed as

@@

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}. Dropout from patients and attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as \emph{missing at random} (MAR), \emph{missing completely at random} (MCAR), and \emph{missing not at random} (MNAR) {[}\protect\hyperlink{ref-weiss2005}{42}{]}. In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest {[}\protect\hyperlink{ref-scheffer2002}{46}{]}. If the data are MCAR, this means that the missingness is completely unrelated to the collected information {[}\protect\hyperlink{ref-potthoff2006}{47}{]}, and in the case of MNAR the missing values are dependent on their value.

An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{48}{]}. In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR {[}\protect\hyperlink{ref-west2014}{40}{]}. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.

@@

To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \eqref{eq:linear-model}), and a LMEM (Equation \eqref{eq:LMEM}) are fitted to each group, using \(\textsf{R}\){[}\protect\hyperlink{ref-r}{38}{]} and the package \emph{nlme}{[}\protect\hyperlink{ref-nlme}{49}{]}.

Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \ref{fig:l-q-response}, A). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \ref{fig:l-q-response}, C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments.

Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect. In other words, Group 1 can be thought as a ``Control'' group and Group 2 as a ``Treatment'' group. From the mean response per group (linear or quadratic), the variability or ``error'' of individual responses within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \ref{fig:l-q-response} B,D).

A more comprehensive exploration of the fit of rm-ANOVA and LMEMs for linear and non-linear longitudinal appears in Figure \ref{fig:linear-cases-Appendix} and Figure \ref{fig:quadratic-cases-Appendix} in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented. We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs. Of notice, Section \ref{longitudinal-GAMs} uses simulated data that does follow reported trends in the biomedical literature.

@@

\caption{Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a LMEM and a rm-ANOVA model. A, D: Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,E: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data but also assigns a global estimate that does not take between-subject variation. C, F: Estimates from the LMEM model in the linear and quadratic case.The LMEM incorporates a random effect for each subject, but this model and the rm-ANOVA model are unable to follow the trend of the data in each group and grossly bias the initial estimates for each group.}\label{fig:l-q-response}
\end{figure}

The simulation shows that the fit produced by the LMEM and the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the ``truth'' of the simulated data (Figure \ref{fig:l-q-response},B, E). When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \ref{fig:l-q-response}, C. Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \ref{fig:l-q-response} A and B) . The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group. Fitting an rm-ANOVA model \eqref{eq:linear-model} or a LMEM \eqref{eq:LMEM} to this data produces the fit that appears in panels E and F in Figure \ref{fig:l-q-response}.

A comparison of the fitted mean response of the LMEM and the rm-ANOVA model to the simulated data in Figure ((\ref{fig:l-q-response}, E, F) indicates that the models are not capturing the changes within each group. Specifically, note that the fitted mean response of both models (panel E, F) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured. The LMEM is only able to account for between-subject variation by providing different intercepts to each subject, but both models are not able to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare panel D with panels E and F). If such a change has important physiological implications,both rm-ANOVA and LMEMs omit it from the fitted mean response. Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

@@

\hypertarget{GAM-theory}{%
\section{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}

@@

Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.

In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible \emph{smooth functions}. This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response. One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference {[}\protect\hyperlink{ref-beck1998}{36}{]}. In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic, and causes bias at the endpoints of the time period.

The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions. A \emph{basis} is a set of functions that spans the mathematical space where the smooths that approximate \(f(x_t \mid \beta_j)\) exist {[}\protect\hyperlink{ref-simpson2018}{34}{]}. For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response {[}\protect\hyperlink{ref-hefley2017}{52}{]}, which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions. In the case of Equation \eqref{eq:GAM}, the basis function is \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.

Commonly used \emph{basis functions} are splines (cubic, thin plate regression among others). A cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data {[}\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}{]}. Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness {[}\protect\hyperlink{ref-wegman1983}{53}{]}. Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure (\ref{fig:l-q-response}, C). The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different basis function will be used. Because there are six timepoints for this group, five knots can be used. The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept). The choice of basis functions is already optimized in the package \emph{mgcv} depending on the number of knots. In Panel A of Figure \ref{fig:basis-plot}, the four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the ``wigliness'' of the fit, each of the basis functions of Panel A is weighted by multiplying it by a coefficient according to the matrix of Panel B. The parameter estimates are penalized where the penalty reduces the ``wigliness'' of the smooth fit to prevent overfitting: A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

@@

\caption{Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2, the intercept basis is not shown. B: Matrix for basis function penalization. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Weighted basis functions. Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B, note the corresponding increase (or decrease) of each basis. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each penalized basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
\end{figure}

@@

The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!}. The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.

@@

The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3, C in {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}. In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
The simulated and real data appear in Figure \ref{fig:sim-smooth-plot}, A and the inlet, respectively.

@@

An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination. In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time. Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in {[}\protect\hyperlink{ref-hefley2017}{52}{]}.

For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment. To estimate this, the model needs to incorporate independent smooths for \emph{Group} and \emph{Day}, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric ( \emph{Day} ) and the other is a factor ( \emph{Group} ). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups. Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:
